---
title: "Data Management and String Handling"
author: Zach del Rosario
date: 2023-04-07
output: github_document
---

# Data Management and String Handling

*Purpose*: This year's DataFest dataset includes a lot of string data. This tutorial is all about working with data, with a particular focus on string data.

**Important Aside**: I'm going to show you *a lot* of tools. **I am not expecting you to memorize everything.** You will have access to this notebook throughout DataFest; if you want to do something like what you see here, you can just *copy and adapt the code* for your own analysis.

## Optional (But Recommended!) Reading

If you have time before DataFest, I highly recommend working through the following reading:

**Working with data**
- *Optional Reading*: [Isolating Data with dplyr](https://rstudio.cloud/learn/primers/2.2) **Note**: In RStudio use `Ctrl + Click` (Mac `Command + Click`) to follow the link.
- *See also*: [Derive Information with dplyr](https://posit.cloud/learn/primers/2.3)

**Finding patterns in strings**
- *Optional Reading*: [RegexOne](https://regexone.com/) **Note**: In RStudio use `Ctrl + Click` (Mac `Command + Click`) to follow the link.
- *Topics*: All lessons in the Interactive Tutorial, Additional Practice Problems are optional. 
- *Note*: The [stringr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf) is a helpful reference for this exercise!

# (LIVE DEMO): Analysis of Famous Speeches

```{r}
library(tidyverse)
```


## Working with data: Fundamentals

### Selecting Rows: `filter()`

### The Pipe

### Deriving values: `mutate()`



## The Gettysburg Address

*Context*

> On June 1, 1865, Senator Charles Sumner referred to the most famous speech ever given by President Abraham Lincoln. In his eulogy on the slain president, he called the Gettysburg Address a "monumental act." He said Lincoln was mistaken that "the world will little note, nor long remember what we say here." Rather, the Bostonian remarked, "The world noted at once what he said, and will never cease to remember it. The battle itself was less important than the speech." (Abraham Lincoln Online)

```{r gettysburg-bliss}
# The full text of the Gettysburg Address (Bliss' copy)
s_gettysburg <- "Four score and seven years ago our fathers brought forth on this continent, a new nation, conceived in Liberty, and dedicated to the proposition that all men are created equal.

Now we are engaged in a great civil war, testing whether that nation, or any nation so conceived and so dedicated, can long endure. We are met on a great battle-field of that war. We have come to dedicate a portion of that field, as a final resting place for those who here gave their lives that that nation might live. It is altogether fitting and proper that we should do this.

But, in a larger sense, we can not dedicate -- we can not consecrate -- we can not hallow -- this ground. The brave men, living and dead, who struggled here, have consecrated it, far above our poor power to add or detract. The world will little note, nor long remember what we say here, but it can never forget what they did here. It is for us the living, rather, to be dedicated here to the unfinished work which they who fought here have thus far so nobly advanced. It is rather for us to be here dedicated to the great task remaining before us -- that from these honored dead we take increased devotion to that cause for which they gave the last full measure of devotion -- that we here highly resolve that these dead shall not have died in vain -- that this nation, under God, shall have a new birth of freedom -- and that government of the people, by the people, for the people, shall not perish from the earth."
```

## Analyzing sentences

### Separate into sentences

The `str_split()` function will split a string on a specified character.

```{r}
str_split("Statement 1, Statement 2", ",")
```

This works as expected when "," is the separating character, but we'll run into a surprise for a ".":

```{r}
str_split(s_gettysburg, ".") %>% 
  .[[1]] %>% 
  .[1:10]
```

The "." character has special meaning; to look for a *literal* period, we have to *escape* it with `\\`.

```{r}
str_split(s_gettysburg, "\\.")
```

This is now something we can fit in a dataset.

```{r}
df_gettysburg_raw <- 
  tibble(
    # Note: We have to extract the nested data using brackets [[1]]
    sentence = str_split(s_gettysburg, "\\.")[[1]]
  )

df_gettysburg_raw 
```

The address is now organized into sentences, but there's a bunch of weird characters. The `\n` characters are *newlines*, a form of *whitespace*. We can remove all the whitespace using `str_trim()`.

```{r}
df_gettysburg <- 
  df_gettysburg_raw %>% 
  mutate(sentence = str_trim(sentence))

df_gettysburg 
```

### Looking for "but"s

```{r}
df_gettysburg %>% 
  mutate(
    but = str_detect(sentence, "but"),
    sentence = str_sub(sentence, end = 20) %>% 
      str_c(., "....")
  ) %>% 
  select(but, sentence)
```

This isn't quite right, we need to catch both upper and lowercase letters.

```{r}
df_gettysburg %>% 
  mutate(
    but = str_detect(sentence, "[Bb]ut"),
    sentence = str_sub(sentence, end = 20) %>% 
      str_c(., "....")
  ) %>% 
  select(but, sentence)
```

### Lincoln's Narrative Index

Now for something surprising! Randy Olson claims that the essence of effective communication can be understood in terms of storytelling. He claims that stories can be understood in terms of three fundamental words: "And", "But", and "Therefore". ABT is a story for another time; for now, we'll focus on the *Narrative Index*, a quantitative tool for assessing how much "story content" there is in a transcript.

*Aside*: For more on the [Narrative Index](http://www.scienceneedsstory.com/blog/the-narrative-index/), check out Olson's website.

The Narrative Index has a very simple definition:

$$NI = \frac{\text{# of `but's in text}}{\text{# of `and's in text}} \times 100$$

Olson claims that a higher NI reflects more "narrative content," and tends to result in more memorable communication. Let's compute the NI for the Gettysburg Address.

The following code computes the Narrative Index for the Gettysburg address:

```{r}
df_ni_gettysburg <- 
  df_gettysburg %>% 
  mutate(
    n_but = str_count(sentence, "[Bb]ut"),
    n_and = str_count(sentence, "[Aa]nd"),
  ) %>% 
  summarize(
    n_but = sum(n_but),
    n_and = sum(n_and),
  ) %>% 
  mutate(ni = n_but / n_and * 100)
df_ni_gettysburg 
```

The NI is ~33 for the Gettysburg Address. But what does that *mean*? For some context, let's take a look at the NI for a few other transcripts.

## The Lincoln-Douglas Debates

```{r}
df_debates <- read_rds("./data/lincoln-douglas-debates.rds")
df_debates
```

The following

```{r}
df_ni_debates <- 
  df_debates %>% 
  mutate(
    n_but = str_count(sentence, "[Bb]ut"),
    n_and = str_count(sentence, "[Aa]nd"),
  ) %>% 
  group_by(speaker, debate) %>% 
  summarize(
    n_but = sum(n_but),
    n_and = sum(n_and),
  ) %>% 
  mutate(ni = n_but / n_and * 100)
df_ni_debates 
```


```{r}
df_ni_debates %>% 
  bind_rows(
    df_ni_gettysburg %>% 
      mutate(
        speaker = "Lincoln",
        debate = "(Gettysburg)"
      )
  ) %>% 
  
  ggplot(aes(debate, ni)) +
  geom_col(aes(fill = speaker), position = "dodge") +
  scale_fill_manual(
    values = c("Douglas" = "blue", "Lincoln" = "red")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 315)) +
  labs(
    x = "Debate",
    y = "Narrative Index\n(# 'but' / # 'and' x 100)"
  )
```

We can make a number of observations:

- Lincoln generally had a higher Narrative Index than Douglas
  - This suggests that Lincoln was following "storytelling" patterns in his debate speeches to a greater degree than Douglas
- The Gettysburg Address had a *much* higher NI than any of these debates
  - This suggests that, even for a "storytelling-heavy" speaker like Lincoln, the Gettysburg Address was a particularly narrative-heavy speech.
  - Given how memorable this speech is, the results seem to agree with Olson's assertion that a high NI corresponds with a more memorable communication.

# (EXERCISES)

