---
title: "Data Management and String Handling: Supplemental Materials"
author: Zach del Rosario
date: 2023-04-07
output: github_document
---

*Purpose*. Process the Lincoln-Douglas debate transcripts into a speaker-labelled, per-sentence format. This will be a useful starting point for the tutorial dataset.

*Note*. Transcripts obtained from the Northern Illinois University: Digital Library, [Lincoln/Net Page](https://digital.lib.niu.edu/illinois/lincoln/topics/lincolndouglas).

# Setup

```{r}
library(tidyverse)

# Data directory
dir_data <- "./data"
# Find all the data files
df_files_all <-
  tibble(filename = list.files(dir_data, full.names = TRUE)) %>%
  mutate(name = str_sub(str_extract(filename, "[^/]+\\."), 1, -2)) %>% 
  filter(str_detect(filename, "\\d+.*\\.txt"))
```

## Load all data

```{r}
df_raw <- 
  df_files_all %>%
  pmap_dfr(
    .,
    function(filename, name) {
      tibble(fulltext = read_file(filename)) %>% 
        mutate(debate = name)
    }
  )

df_raw
```

## Separate sentences


```{r}
df_sentences <- 
  df_raw %>% 
  mutate(sentence = str_split(fulltext, "(?<!Mr|Ms|Mrs)\\.")) %>% 
  select(-fulltext) %>% 
  unnest_longer(sentence)

df_sentences
```

## Label Speaker

The debates took a consistent format, which is reflected in the transcript annotations. We can use this to help label the speaker for each sentence.

I've copied the headers from the text. Let's inspect them for patterns we can use:

### Headers

01-ottawa
- `Mr. Douglas's Opening Speech.`
- `Mr. Lincoln's Reply in the Ottawa Joint Debate.`
- `Mr. Douglas's Rejoinder in the Ottawa Joint Debate.`
02-freeport
- `Mr. Lincoln's Opening Speech.`
- `Mr. Douglas's Reply in the Freeport Joint Debate.`
- `Mr. Lincoln's Rejoinder in the Freeport Joint Debate.`
03-jonesboro
- `Mr. Douglas's Opening Speech`
- `Mr. Lincoln's Reply in the Jonesboro Joint Debate`
- `Mr. Douglas's Rejoinder in the Jonesboro Joint Debate.`
04-charleston
- `Mr. Lincoln's Opening Speech`
- `Senator Douglas's Reply in the Charleston Joint Debate`
- `Lincoln's Rejoinder in the Charleston Joint Debate`
05-galesburg
- `Mr. Douglas's Opening Speech`
- `Mr. Lincoln's Reply at Galesburg Joint Debate`
- `Mr. Douglas's Rejoinder in the Galesburgh Joint Debate`
06-quincy
- `Mr. Lincoln's Opening Speech`
- `Mr. Douglas's Reply in the Quincy Joint Debate`
- `Mr. Lincoln's Rejoinder in the Quincy Joint Debate`
07-alton
- `Senator Douglas's Opening Speech.`
- `Mr. Lincoln's Reply in the Alton Joint Debate.`
- `Mr. Douglas's Rejoinder in the Alton Joint Debate.`

*Observations*
- Douglas is alternatively referred to as `Senator Douglas` or `Mr. Douglas`; we'll have to ignore his title.
- All debates followed the pattern of `Opening Speech`, `Reply`, then `Rejoinder`. We can use this consistent pattern to help label the speaker for each sentence.

```{r}
df_speaker <- 
  df_sentences %>% 
  extract(
    sentence,
    c("speaker", "part"),
    "(?<name>Lincoln|Douglas).*(?<part>Opening Speech|Reply|Rejoinder)",
    remove = FALSE
  ) %>% 
  fill(speaker, part) %>% 
  select(debate, speaker, part, sentence)

df_speaker
```

## Write the data

```{r}
write_rds(df_speaker, "./data/lincoln-douglas-debates.rds")
```


# Analyze

## Narrative Index

Randy Olson proposed the [Narrative Index](http://www.scienceneedsstory.com/blog/the-narrative-index/) as a quantitative way to summarize the amount of "narrative content" of a transcript. The narrative index is defined as

$$NI = \frac{\text{# of `but's in text}}{\text{# of `and's in text}} \times 100$$

The following code computes the narrative index for each debate transcript, grouped by speaker.

```{r ni-compute}
df_ni <- 
  df_speaker %>% 
  mutate(
    n_but = str_count(sentence, "[Bb]ut"),
    n_and = str_count(sentence, "[Aa]nd"),
  ) %>% 
  group_by(debate, speaker) %>% 
  summarize(
    n_but = sum(n_but),
    n_and = sum(n_and)
  ) %>% 
  mutate(ni = n_but / n_and * 100) %>% 
  ungroup()

df_ni
```

Visualize the results to compare against Olson's work.

```{r ni-viz}
df_ni %>% 
  ggplot(aes(debate, ni)) +
  geom_col(aes(fill = speaker), position = "dodge") +
  scale_fill_manual(
    values = c("Douglas" = "blue", "Lincoln" = "red")
  ) +
  theme_minimal() +
  labs(
    x = "Debate",
    y = "Narrative Index\n(# 'but' / # 'and' x 100)"
  )
```

This figure is very close to Olson's results, but my NI is off Olson's results by a bit for many cases. For instance, Olson's figure suggests that Lincoln had NI = 20 for the Alton debate. However, I found ~19 instead. Perhaps he used a different source for his transcripts?
